---
title: 'Inference for categorical data'
output:
  html_document:
    css: ./lab.css
    highlight: pygments
    theme: cerulean
  pdf_document: default
---

1.  In the first paragraph, several key findings are reported. Do these 
    percentages appear to be *sample statistics* (derived from the data 
    sample) or *population parameters*?  
    
    <b>They are sample statistics.</b>  
    

2.  The title of the report is "Global Index of Religiosity and Atheism". To
    generalize the report's findings to the global human population, what must 
    we assume about the sampling method? Does that seem like a reasonable 
    assumption?  
    
    <b>To generalize the report we must assume the sampling method is random and independent and the sample is sufficiently large; in this case it is a reasonable assumption to make.</b>  

```{r head-data, eval=TRUE}
load("more/atheism.RData")
```

3.  What does each row of Table 6 correspond to? What does each row of 
    `atheism` correspond to?  
    
    <b>Each row of table six represents a country. Each row in our dataframe, 'atheism', represents an individual participant.</b>  
    

4.  Using the command below, create a new dataframe called `us12` that contains
    only the rows in `atheism` associated with respondents to the 2012 survey 
    from the United States. Next, calculate the proportion of atheist 
    responses. Does it agree with the percentage in Table 6? If not, why?

```{r us-atheism, eval=TRUE}
us12 <- subset(atheism, nationality == "United States" & year == "2012")
propAtheist<-sum(us12$response == 'atheist')/nrow(us12)
cat("The proportion of American atheists in our dataset is",round(propAtheist*100,0),"%, which confirms the total of confirmed atheists in Table 6.")
```

5.  Write out the conditions for inference to construct a 95% confidence
    interval for the proportion of atheists in the United States in 2012.
    Are you confident all conditions are met?  
    
    <b>The sample is randomly selected.  
    Each individual is an independent 'trial' (atheist vs non-atheist).  
    To satisfy success-failure criterion:  
```{r message=FALSE,warning=FALSE}
#US, p & n from Table 6
ath <- .05*1002
nath <- (1-.05)*1002
cat("Since",ath,"&",nath,"are sufficiently large, the success-failure criterion is met.")
```


```{r us-atheism-ci, eval=TRUE, tidy = FALSE}
inference(us12$response, est = "proportion", type = "ci", method = "theoretical", 
          success = "atheist")
```


6.  Based on the R output, what is the margin of error for the estimate of the 
    proportion of the proportion of atheists in US in 2012?  
```{r message=FALSE,warning=FALSE}
#for confidence interval 95%, z-score is 1.96; SE is .0069 from above output

ME <- 1.96*.0069
cat("The margin of error is",ME)

```

7.  Using the `inference` function, calculate confidence intervals for the 
    proportion of atheists in 2012 in two other countries of your choice, and 
    report the associated margins of error. Be sure to note whether the 
    conditions for inference are met. It may be helpful to create new data 
    sets for each of the two countries first, and then use these data sets in 
    the `inference` function to construct the confidence intervals.  
```{r message=FALSE,warning=FALSE}
#Argentina and Australia, subsets
Arg12 <- subset(atheism, nationality == "Argentina" & year == "2012")
Aus12 <- subset(atheism, nationality == "Australia" & year == "2012")
```  

<b>Argentina</b>  


```{r message=FALSE,warning=FALSE}

propArgAtheist<-sum(Arg12$response == 'atheist')/nrow(Arg12)
cat("The proportion of Argentinian atheists in our dataset is",round(propArgAtheist*100,0),"%, which confirms the total of confirmed atheists in Table 6.")

```  

```{r message=FALSE,warning=FALSE}
#Argentina, p & n from Table 6
ath <- .07*1002
nath <- (1-.07)*1002
cat("Argentina: since",ath,"&",nath,"are sufficiently large, the success-failure criterion is met.")
```


```{r message=FALSE,warning=FALSE}

#inference, Argentina
inference(Arg12$response, est = "proportion", type = "ci", method = "theoretical", 
          success = "atheist")

```  

```{r message=FALSE,warning=FALSE}
#Argentina: for confidence interval 95%, z-score is 1.96; SE is .0081 from above output

MEarg <- 1.96*.0081
cat("Argentina: the margin of error is",MEarg)

```  


<b>AUSTRALIA</b>  


```{r message=FALSE,warning=FALSE}
#Australia
propAusAtheist<-sum(Aus12$response == 'atheist')/nrow(Aus12)
cat("The proportion of Australian atheists in our dataset is",round(propAusAtheist*100,0),"%, which confirms the total of confirmed atheists in Table 6.")
```


```{r message=FALSE,warning=FALSE}
#Australia, p from text & n from Table 6
ath <- .10*1040
nath <- (1-.10)*1040
cat("Australia: since",ath,"&",nath,"are sufficiently large, the success-failure criterion is met.")
```


```{r message=FALSE,warning=FALSE}

#inference, Australia
inference(Aus12$response, est = "proportion", type = "ci", method = "theoretical", 
          success = "atheist")

```  

```{r message=FALSE,warning=FALSE}
#Australia: for confidence interval 95%, z-score is 1.96; SE is .0093 from above output

MEaus <- 1.96*.0093
cat("Australia: the margin of error is",MEaus)

```



```{r me-plot, eval=TRUE}
n <- 1000
p <- seq(0, 1, 0.01)
me <- 2 * sqrt(p * (1 - p)/n)
plot(me ~ p, ylab = "Margin of Error", xlab = "Population Proportion")
```

8.  Describe the relationship between `p` and `me`.  

<b>We see that `me` is at a maximum when `p` = .5; as `p` varies to 1 or 0, `me` decreases.</b>  



```{r sim-np, eval=TRUE}
p <- 0.1
n <- 1040
p_hats <- rep(0, 5000)

for(i in 1:5000){
  samp <- sample(c("atheist", "non_atheist"), n, replace = TRUE, prob = c(p, 1-p))
  p_hats[i] <- sum(samp == "atheist")/n
}

hist(p_hats, main = "p = 0.1, n = 1040", xlim = c(0, 0.18))
```


9.  Describe the sampling distribution of sample proportions at $n = 1040$ and 
    $p = 0.1$. Be sure to note the center, spread, and shape.\
    *Hint:* Remember that R has functions such as `mean` to calculate summary
    statistics.  
```{r message=FALSE,warning=FALSE}
summary(p_hats)
IQR(p_hats)
```  
<b>The distribution is normal, unimodal, centered at 0.1, with mean and median approximately equal (.09969 and .09904 respectively), and IQR `r IQR(p_hats)`. 

10. Repeat the above simulation three more times but with modified sample
    sizes and proportions: for $n = 400$ and $p = 0.1$, $n = 1040$ and 
    $p = 0.02$, and $n = 400$ and $p = 0.02$. Plot all four histograms
    together by running the `par(mfrow = c(2, 2))` command before creating the 
    histograms. You may need to expand the plot window to accommodate the 
    larger two-by-two plot. Describe the three new sampling distributions. 
    Based on these limited plots, how does $n$ appear to affect the 
    distribution of $\hat{p}$? How does $p$ affect the sampling distribution?  
    
```{r message=FALSE, warning = FALSE}

p <- 0.1
n <- 400
p_hats400.1 <- rep(0, 5000)

for(i in 1:5000){
  samp <- sample(c("atheist", "non_atheist"), n, replace = TRUE, prob = c(p, 1-p))
  p_hats400.1[i] <- sum(samp == "atheist")/n
}


p <- 0.02
n <- 1040
p_hats1040.02 <- rep(0, 5000)

for(i in 1:5000){
  samp <- sample(c("atheist", "non_atheist"), n, replace = TRUE, prob = c(p, 1-p))
  p_hats1040.02[i] <- sum(samp == "atheist")/n
}


p <- 0.02
n <- 400
p_hats400.02 <- rep(0, 5000)

for(i in 1:5000){
  samp <- sample(c("atheist", "non_atheist"), n, replace = TRUE, prob = c(p, 1-p))
  p_hats400.02[i] <- sum(samp == "atheist")/n
}

par(mfrow = c(2, 2))

hist(p_hats, main = "p = 0.1, n = 1040", xlim = c(0, 0.18))
hist(p_hats400.1, main = "p = 0.1, n = 400", xlim = c(0, 0.18))
hist(p_hats1040.02, main = "p = 0.02, n = 1040", xlim = c(0, 0.18))
hist(p_hats400.02, main = "p = 0.02, n = 400", xlim = c(0, 0.18))
    
```

<b>As expected, an increase in sample size constricts spread, while a decrease in probability simultaneously shifts the center toward 0 and constricts the spread of the distribution.</b>  

```{r message=FALSE,warning=FALSE}

par(mfrow = c(1, 1))

```

11. If you refer to Table 6, you'll find that Australia has a sample 
    proportion of 0.1 on a sample size of 1040, and that Ecuador has a sample 
    proportion of 0.02 on 400 subjects. Let's suppose for this exercise that 
    these point estimates are actually the truth. Then given the shape of 
    their respective sampling distributions, do you think it is sensible to 
    proceed with inference and report margin of errors, as the reports does?  
    
    <b>Australia satifies the conditions of inference, as show above, and repeated below.</b>  
    
```{r message=FALSE,warning=FALSE}
#Australia, p & n from Table 6
ath <- .10*1039
nath <- (1-.10)*1039
cat("Australia: since",ath,"&",nath,"are sufficiently large, the success-failure criterion is met.")
```  

Checking Ecuador yields:  

```{r message=FALSE,warning=FALSE}
#Ecuador, p & n from Table 6
ath <- .02*400
nath <- (1-.02)*400
cat("Since",ath,"&",nath,"are not sufficiently large, the success-failure criterion is not met for Ecuador.")
``` 
<b>Since the value of 8 calculated for Ecuador is close to 10, we can allow a generalization for the population at large with a responsibly delivered caveat; otherwise inference is not sensible and more data need to be collected.</b>  

* * *
## On your own

The question of atheism was asked by WIN-Gallup International in a similar 
survey that was conducted in 2005. (We assume here that sample sizes have 
remained the same.) Table 4 on page 13 of the report summarizes survey results 
from 2005 and 2012 for 39 countries.

-   Answer the following two questions using the `inference` function. As 
    always, write out the hypotheses for any tests you conduct and outline the
    status of the conditions for inference.

    **a.** Is there convincing evidence that Spain has seen a change in its 
    atheism index between 2005 and 2012?\
    *Hint:* Create a new data set for respondents from Spain. Form
    confidence intervals for the true proportion of athiests in both
    years, and determine whether they overlap.  
    
<b>$H_0$: no convincing evidence that difference exists in the atheism indexes  
$H_A$: there is convincing evidence that difference exists in the atheism indexes</b>  

```{r message=FALSE,warning=FALSE}
#Spain, subsets for 2005 and 2012
Esp05 <- subset(atheism, nationality == "Spain" & year == 2005)
Esp12 <- subset(atheism, nationality == "Spain" & year == 2012)

propEsp05Atheist<-sum(Esp05$response == 'atheist')/nrow(Esp05)
propEsp12Atheist<-sum(Esp12$response == 'atheist')/nrow(Esp12)

cat("The proportion of Spanish atheists in our dataset by year are 2005:",round(propEsp05Atheist*100,0),"% and 2012:",round(propEsp12Atheist*100,0),"%, which approximates the total of confirmed atheists in Table 6.\n\n")

#Spain, p & n from Tables in article
esp05.ath <- .1*1146
esp05.nath <- (1-.1)*1146
cat("Since",esp05.ath,"&",esp05.nath,"are sufficiently large, the success-failure criterion is met for Spain in 2005.\n\n")

esp12.ath <- .09*1146
esp12.nath <- (1-.09)*1146
cat("Since",esp12.ath,"&",esp12.nath,"are sufficiently large, the success-failure criterion is met for Spain in 2012.\n\n")

#inference, Spain 2005
inference(Esp05$response, est = "proportion", type = "ci", method = "theoretical", 
          success = "atheist")

#inference, Spain 2012
inference(Esp12$response, est = "proportion", type = "ci", method = "theoretical", 
          success = "atheist")

```  

<b>We see that the confidence intervals overlap; therefore, we accept the null hypothesis, and assert that there is no convincing evidence that a change exists in the atheism indexes of Spain between 2005 and 2012.</b>  


    **b.** Is there convincing evidence that the United States has seen a
    change in its atheism index between 2005 and 2012?  
    
<b>$H_0$: no convincing evidence that difference exists in the atheism indexes  
$H_A$: there is convincing evidence that difference exists in the atheism indexes</b>  

```{r message=FALSE,warning=FALSE}

#US, subsets for 2005 and 2012
US05 <- subset(atheism, nationality == "United States" & year == 2005)
US12 <- subset(atheism, nationality == "United States" & year == 2012)

propUS05Atheist<-sum(US05$response == 'atheist')/nrow(US05)
propUS12Atheist<-sum(US12$response == 'atheist')/nrow(US12)

cat("The proportion of American atheists in our dataset by year are 2005:",round(propUS05Atheist*100,0),"% and 2012:",round(propUS12Atheist*100,0),"%, which approximates the total of confirmed atheists in Table 6.\n\n")

#US, p & n from Tables in article
US05.ath <- .01*1002
US05.nath <- (1-.01)*1002
cat("Since",US05.ath,"&",US05.nath,"are sufficiently large, the success-failure criterion is met for America in 2005.\n\n")

US12.ath <- .05*1002
US12.nath <- (1-.05)*1002
cat("Since",US12.ath,"&",US12.nath,"are sufficiently large, the success-failure criterion is met for America in 2012.\n\n")

#inference, US 2005
inference(US05$response, est = "proportion", type = "ci", method = "theoretical", 
          success = "atheist")

#inference, US 2012
inference(US12$response, est = "proportion", type = "ci", method = "theoretical", 
          success = "atheist")

```  

<b>We see that the confidence intervals do not overlap; therefore, we reject the null hypothesis, and assert that there is convincing evidence that a change exists in the atheism indexes of the United States between 2005 and 2012.</b>  
    

-   If in fact there has been no change in the atheism index in the countries 
    listed in Table 4, in how many of those countries would you expect to 
    detect a change (at a significance level of 0.05) simply by chance?\
    *Hint:* Look in the textbook index under Type 1 error.  
    
    <b>With significance level of .05 we would expect to detect a change in $.05 \times 39 \text{countries} \approx$ `r round(.05 * 39,0)'</b>  
    

-   Suppose you're hired by the local government to estimate the proportion of 
    residents that attend a religious service on a weekly basis. According to 
    the guidelines, the estimate must have a margin of error no greater than 
    1% with 95% confidence. You have no idea what to expect for $p$. How many 
    people would you have to sample to ensure that you are within the 
    guidelines?\
    *Hint:* Refer to your plot of the relationship between $p$ and margin of 
    error. Do not use the data set to answer this question.  
    
$$ ME = z \times SE\implies ME=z \times\sqrt{\bigg(\frac{p(1-p)}{n}\bigg)}\implies n=\frac{p\times(1-p)\times z^2}{ME^2}$$  
    
```{r message=FALSE,warning=FALSE}    

#ME is at a maximum with p = .5, from above
#z-score for 95% confidence interval, from text
#margin of error set to .01, from text

n <- (.5*.5*(1.96^2))/(.01^2)
cat("We would have to sample",ceiling(n),"people to remain within the guidelines.")
```  


